---
title: "Task 1 Question 2"
author: "Meredith Gavin"
date: "7/14/2023"
output: pdf_document
---

# PREP WORK

## Libraries

```{r}
library(haven)
library(dplyr)
library(tidyverse)
library(magrittr)
library(readr)
library(ggplot2)
library(lubridate)
```

## Set Working Directory

```{r}
setwd("~/Desktop/pre_doc_task_1")
```


## Load Data

```{r}
ma_enrollment <- read_csv("scp_1205.csv", col_names = FALSE)
```

## QUESTION 2: DATA MANIPULATION

### Rename Headers

```{r}
ma_enrollment <- ma_enrollment %>%
  rename(countyname = X1,
         state = X2,
         healthplanname = X3,
         typeofplan = X4,
         countyssa = X5,
         eligibles = X6,
         enrollees = X7,
         penetration = X8,
         ABrate = X9)
```



### Change NAs to 0 for  eligibles, enrollees, and penetration variables and remove Purto Rico and Guam

```{r}
ma_enrollment <- ma_enrollment %>%
  mutate_at(c("eligibles", "enrollees", "penetration"),
            as.double) %>% # make variables numeric
  mutate_at(c("eligibles", "enrollees", "penetration"),
            ~replace_na(., 0)) %>% # change NAs to zeroes
  filter(state != "GU" & state != "PR") # filter out Puerto Rico and Guam
```

### 

```{r}
# create numberofplans1 variable
ma_enrollment_numberofplans1 <- ma_enrollment %>%
  filter(enrollees > 10) %>%
  group_by(state, countyname) %>%
  summarise(numberofplans1 = n()) %>%
  ungroup()

```

```{r}
#create numberofplans2 variable

ma_enrollment_numberofplans2 <- 
  ma_enrollment %>% 
  filter(penetration > 0.5) %>%
  group_by(state, countyname) %>%
  summarise(numberofplans2 = n())
```

### Merge new variables to master dataframe

```{r}
ma_enrollment <- ma_enrollment %>% 
  left_join(ma_enrollment_numberofplans1, 
            by = c("state", "countyname")) %>%
  left_join(ma_enrollment_numberofplans2, 
            by = c("state", "countyname"))
```

### create totalenrollees variable

```{r}
county_enrollment <- ma_enrollment %>%
  group_by(state, countyname) %>%
  summarise(totalenrollees = sum(enrollees)) %>%
  ungroup()
```



### merge totalenrollees to master df and create totalpenetration 

```{r}
ma_enrollment_final <- ma_enrollment %>%
  left_join(county_enrollment,
            by = c("state", "countyname")) %>%
  mutate(totalpenetration = (totalenrollees / eligibles) * 100) %>%
  select(
    countyname,
    state,
    numberofplans1,
    numberofplans2,
    countyssa,
    eligibles,
    totalenrollees,
    totalpenetration
  ) %>%
  arrange(state, countyname)
```

### save final dataframe as csv

```{r}
write_csv(ma_enrollment_final, "ma_enrollment_final.csv")
```












